Titeln på avhandlingens språk

### 2022-11-20 body/part1/overview.tex
structuring

This thesis presents seven main key contributions to the literature, which we summarize below:

@@

\todo[inline]{Intro and conclusions from EdgeDroid 1.0 paper}

There is increasing interest from academia and industry in novel applications such as immersive augmented reality (AR) or wearable cognitive assistance (WCA)~\cite{Chatzopoulos:Hyperion,Ha:TowardsWearableCogAssist}, also known as human-in-the-loop applications.
These applications aim to seamlessly integrate into the lives of users to provide real-time, context-aware information by capturing user and environment information and leveraging compute-intensive algorithms to analyze the data in order to provide real-time feedback to the user.
Sensory input, such as video, audio, and other user-related data such as orientation and movement, are examples of what is typically captured, while the backend generally employs machine learning technologies such as Deep Neural Networks (DNN)~\cite{Ha:TowardsWearableCogAssist}.
\Cref{fig:pipeline2} depicts such a system.
These applications are highly latency sensitive, measuring latency as the time from the capture of the sensory information until feedback is received.
Delays above a certain threshold can hurt the user experience or even make the application unusable~\cite{Chen:AnEmpiricalStudyOfLatency}.

In literature, these challenging latency requirements have so far mainly been addressed through research on the optimal placement of the compute process.
There is a broad understanding today that with the advent of edge computing, human-in-the-loop applications will become viable~\cite{Bittmann_Edge,flinn2012cyber,Chen:AnEmpiricalStudyOfLatency,Ha:JITProvisioning}.
However, with respect to end-to-end latency, there are many more trade-offs involved than merely the question of where the compute backend is placed.
A human-in-the-loop application consists of various processing steps that can be influenced during the development of the application.
What kind of compression to apply to the sensory input on the uplink; which backend algorithms to utilize; how to stage the backend; when to send feedback to the human users; and how to manage congestion on the loop, as well as wireless channel fluctuations --- all these design choices impact the latency of the application.
There are also many design choices in the infrastructure: how is the sensory input conveyed to the point of computation (i.e.\ by which wireless system; with which transmission/prioritization scheme); which hardware is running at the backend; which operating system; how is the feedback conveyed back to the human user?
Finally, in production use these applications will most likely run concurrently with others.
How does this, together with other best-effort applications, impact the latencies perceived by the human user?
Existing studies~\cite{Ha:TowardsWearableCogAssist, Chen:EarlyImplementation, satya2009case,Chatzopoulos:Hyperion} of this class of applications have only lightly touched upon these issues~\cite{Chen:AnEmpiricalStudyOfLatency}.
On the other hand, recently published models for end-to-end latency of edge computing architectures~\cite{Zubaidy15,Schiessl17} are quite complex, while not accounting for the specifics of human-in-the-loop applications.
We only have a coarse understanding of the many degrees of freedom upon which end-to-end latency depends.

@@

Benchmarking human-in-the-loop applications is hard, given their tight interaction with human users who complicate the scaling and repeatability of experiments.
In this paper, we have presented a benchmarking suite for this type of applications, called EdgeDroid 1.0, capable of cutting out the need for users in performance evaluations.
We achieve this by employing pre-recorded sensory input traces which we play over the network to the real application backend, employing a parameterized user model to react to feedback.
We demonstrate its utility through a series of use case scenarios, from which we are able to extract metrics regarding latency both in regards to the application itself and the hardware stack.
We believe the EdgeDroid 1.0 suite thus represents an important first step towards enabling inexpensive and low-complexity large-scale research on the scaling limits of this type of applications, a requirement for wide adoption of the technology.

Nonetheless, there is still future work to be done.

The user model presented in this paper is only preliminary, and we are currently conducting research in characterizing user behavior when interacting with WCA applications in order to present a more complete model in the future.
As mentioned in \Cref{sec:implementation}, in EdgeDroid 1.0, our model is that of a user who does not suffer any of the shortcomings of real human users such as annoyance, fatigue, frustration, nausea.
Rather, EdgeDroid 1.0 models a perfectly stoic user who is like an automaton and responds in a precisely reproducible and deterministic manner to the same system stimulus every time.
Of course, no real human user is an automaton.
In the future, we envision creating  many versions of EdgeDroid (i.e., EdgeDroid 2.0, EdgeDroid 3.0, etc.) that embody more human-like user models that more accurately emulate attributes such as those mentioned above.
Experimental validation of these human-like user models via user studies will be an important part of our future work.

We are also working on expanding the benchmarking suite to also work first with other types of Wearable Cognitive Assistance, and later with other categories of human-in-the-loop applications.
Other types of WCA we will consider in future iterations of the tool include real-time task-assistance WCA applications (such as the Ping-Pong application described in~\cite{Chen:EarlyImplementation}), which don't have a linear task model like task-guidance WCA and have tighter latency bounds and context- and information-providing WCA applications, for instance, applications which recognize faces and provide relevant social-media information related to that person.
The latter also do not have a linear task model, but present more lax latency bounds.

\todo[inline]{Intro and conclusions from CLEAVE paper}

The number and applications of \acp{CPS}~\cite{Rajkumar2010CPS} --- i.e.\ systems in which a real, physical mechanism is controlled by a computer --- have exploded in recent years.
However, this rapid increase in adoption has mostly been limited to industrial contexts.
Although \acp{CPS} present huge opportunities for all facets of society, they have yet to reach our daily lives in any relevant scale due to their stringent operational requirements.
This is about to change, however, as with the advent of novel wireless communication technologies as well as networking paradigms, such as cellular 5G and edge computing~\cite{Satya2017Emergence}, consumer-grade \acp{CPS} will be made possible.
These technologies meet two key requirements of \acp{CPS}: real-time capabilities (through extremely low end-to-end latencies), and context- and locality-awareness, and will most likely become the backbone of \ac{CPS} in the future.

\acp{NCS}~\cite{Gupta2010NCSOverview}, a type of \ac{CPS} wherein multiple networked actuators and sensors form a part of the same automatic control system will benefit from the adoption of these technologies.
Depending on the physical system being controlled, \acp{NCS} can have stringent timing and reliability requirements for communication that conventional cloud paradigms and cellular networks cannot meet~\cite{Wan2020Efficient}.
This necessitates sophisticated tools for the performance evaluation of future system architectures, as well as novel NCS design paradigm.

%Due to their potential advantages for industrial and commercial settings, there exist works~\cite{Zhang2016Survey} dedicated to the modelling and performance characterization of \acp{NCS}, improving NCSs by distributing control functions across networks, facilitating centralized coordination, control, and monitoring.

One the one hand, related literature in \acp{NCS} leverages to a large extent theoretical models, at the price of being able to capture networked systems effects only on a coarse level. 
%follows a theoretical approach, and only a small fraction of it deals with experimental studies.
On the other, there exist several approaches when considering experimental methodologies. 
%A number of works concerning \acp{NCS} deal with experimental studies.
%\acp{NCS} have an inherently inter-domain nature intertwining knowledge from the fields of communications, computing, and control theory in ways that cannot be studied in isolation, leading to various different approaches to such studies.
One such approach uses setups in which the complete system is built on top of real hardware.
This approach is employed in the works of Baumann \emph{et al.}~\cite{Baumann2018LowPower} and Cuenca \emph{et al.}~\cite{Cuenca2019UAV}; in both of these, the authors implement their approach on physical testbeds.
Conversely, other studies choose to instead use completely \emph{simulated} \ac{NCS} setups.
The authors in\ \cite{Ma2019DynamicSched} have opted for such an approach.
These studies often employ combinations of physical and network simulation tools trying to capture the complex dynamics of \acp{NCS}.
Finally, some experimental studies instead employ \emph{virtualized} approaches, in which either
\begin{enumerate*}[itemjoin={{; }}, itemjoin*={{; or }}]
    \item a real network interacts with a simulated or emulated control system~\cite{Wang2020VoltageControl}
    \item a simulated network interacts with a real control system~\cite{Natale2004InvPendEthernet}.
\end{enumerate*}

As evidenced above, experimental research in \acp{NCS} includes varied heterogeneous hardware and software platforms, methodologies and key performance indicators.
This, in turn, leads to hardware, software, and methodology fragmentation, as different studies tend to prefer approaches more favored in their respective communities.
Furthermore, existing studies tend to focus on individual aspects and components of a system, thus producing results which do not provide a complete image of the \ac{NCS}.
This has caused a gap in knowledge pertaining to the reproducibility and comparison of experimental studies on these systems.

Zoppi \emph{et al.}~\cite{Zoppi2020NCSBench} made the first (and to the best of our knowledge, the only) attempt at tackling this challenge in their work.
In their work, they proposed a platform called NCSBench, to be used for reproducible benchmarking in NCS.\ 
Their methodology utilizes joint knowledge of control, computation, and communication. 
In their work various architectural elements and the corresponding delays associated with the NCS are modelled. 
Multiple experimental parameters and certain observable key performance indicators are defined and utilized in the implementation. 
This work however utilizes a physical LEGO\textregistered{}\ Mindstorms EV3 Core Set\texttrademark{}\  based plant for the implementation, preventing instantaneous changes in plant characteristics and component parametrizations.
Furthermore, relying on physical objects like an inverted pendulum limits scalability of the experimentation in practice.

The rest of this paper is structured as follows.
\cref{sec:approach} presents the design principles and implementation of the framework.
In \cref{sec:experiments}, we present a use case scenario which validates the utility, flexibility, and repeatability of \ac{CLEAVE}.
Finally, in \cref{sec:conclusion} we conclude this work.

The issue of repeatable and scalable benchmarks has been largely glossed over in \ac{NCS} literature, as existing experimental research studies tend to implement \emph{ad-hoc} solutions.

There is still, however, work to be done.
We are extending the number of plant and controller implementations on the framework, with the goal of creating an open library of \acp{NCS} to share with the community.
At the moment, the interactions of \ac{CLEAVE} and tools such as Docker are still quite superficial.
Our goal is to achieve a much tighter integration, e.g.\ by providing the toolkit as pre-packaged container images.
Finally, the validity of the results obtained by the framework will have to be verified through more thorough, realistic scenarios than what we have been able to show in this work.
In particular, we intend to perform large-scale experimentation targetting 5G cellular deployments, as this technology is set to become the backbone of edge networks in the near future.

\todo[inline]{Intro and conclusions from Ainur paper}

Experimental research in the area of wireless networking has received ever increasing attention over the last years, driven, on the one hand, by the complexity of modern networked systems and corresponding applications. 
On the other, networked systems are more and more based on software instead of dedicated hardware, which allows experimental testbeds to be rededicated simply through an update as system versions evolve --- in contrast to the redeployment of hardware necessitated \numrange[range-phrase={--}]{10}{15} years ago.
The complexity of these systems, as well as their softwarization are expected to continue growing, driving in turn an expanding interest in testbed-based experimental research in wireless systems.

Over the last years, several small- to mid-scale testbeds have emerged that leverage a large degree of freedom with respect to hardware and software, for example, the 
\begin{enumerate*}[itemjoin={{, }}, itemjoin*={{, and }}]
    \item \acs*{COSMOS}
    \item \acs*{POWDER}
    \item \acs*{ARA}
    \item Drexel Grid \ac{SDR}
\end{enumerate*} testbeds.
\acs{COSMOS}~(\emph{\acl{COSMOS}})\acused{COSMOS} is a testbed spanning an area of roughly \num{1} square mile (\SI{2.6}{\kilo\meter\squared}) featuring \acp{SDR}, \si{\milli\meter}-wave equipment, optical fibers, cloud integration, and compute for core network functionality and application data processing~\cite{Cosmos1,cosmos2}.
It contains over \num{200} rooftop, intermediate, and mobile nodes, and is controlled and managed by a central node.
\ac{COSMOS} relies on the \ac{OMF} (originally developed for ORBIT~\cite{orbit}), and employs the \ac{OEDL}, a domain-specific imperative language based on Ruby, for experiment development and definition.

\acs{POWDER}~(\emph{\acl{POWDER}})\acused{POWDER} promises research on wireless and mobile networks with a level of programmability down to the waveform~\cite{powder}.
The testbed spans a \SI{15}{\kilo\meter\squared} area and features about \num{15} fixed programmable radio nodes, based on off-the-shelves \acp{SDR} and featuring edge-like compute capabilities and integration with cloud resources, which interact with \num{50} mobile nodes. 
\ac{POWDER} experiments are defined and developed in \emph{profiles}, which correspond to \ac{VM} images containing the necessary software and configurations.
These profiles are defined through using \ac{RSpec}\footnote{\url{https://groups.geni.net/geni/wiki/GENIExperimenter/RSpecs}} documents.

The \ac{ARA} platform is an at-scale testbed for wireless research spanning a rural area with a diameter of over \SI{60}{\kilo\meter} in Iowa~\cite{zhang2022ara}.
Its core goal is the study and deployment of advanced wireless platforms and technologies in real-world agricultural and rural settings.
It includes a broad range of wireless platforms ranging from low-\ac{UHF} massive \ac{MIMO} to \unit{\milli\meter}Wave, deployed through both \acp{SDR} and programmable \ac{COTS} radios, as well as automated ground vehicles, cameras and sensors.
\ac{ARA}'s software stack, ARASoft, is based upon the highly flexible and powerful \ac{CHI} software suite from the Chameleon testbed project~\cite{keahey2020lessons}, which in turn is based on the widely adopted OpenStack~\cite{openstack} cloud-computing framework.
This affords the \ac{ARA} testbed a high degree of flexibility, as well as lowers the learning curve for new contributors and users.

Finally, the \emph{Drexel Grid \ac{SDR} Testbed} features \acp{SDR} that connect either over-the-air, through a channel emulator, or over a combination of the two, to facilitate realistic and reproducible experimentation~\cite{DrexelGrid}.
Primarily intended for \ac{SDR}-centric research, it does not integrate any core, cloud or edge components.
However, the testbed extensively employs the \ac{LXC} runtime for the deployment of both experimental code and \ac{SDR} software, which affords users great freedom when it comes to the development of experiments.

Experimentation is key to to fully understanding the implications of next-generation wireless systems, cloud, and edge computing paradigms, and thus more of these testbeds are sure to emerge in the near future.
Yet, little work has so far been devoted to general-purpose, hardware-agnostic software frameworks for the management and automation of such systems.
Existing platforms implement their own, ad-hoc software solutions which are not compatible with other testbeds, and in many cases are not even compatible with reigning cloud-native standards.
This is, for instance, the case with \ac{COSMOS} and \ac{POWDER}; their reliance on domain-specific languages limits their integration with cloud-native solutions, which generally build upon general-purpose languages such as Python and Go.
These testbeds further leverage virtualization technology based on \acp{VM} instead of more lightweight and edge-compatible solutions such as containers.

To the best of our knowledge, CloudRAFT~\cite{cloudraft} is the only work to tackle (to a certain extent) this challenge.
CloudRAFT corresponds to a cloud-based framework for mobile network experimentation, with a focus on simplifying the management of testbed resources.
The goal of this project is to integrate, coordinate, share, and improve upon existing testbeds; and employs pre-built \acp{VM} containing the necessary software for experiments.
Although it provides some automation for testbed resource provisioning and experiment execution, its focus is largely rather on the sharing and partitioning of testbed systems.
Testbeds currently working with CloudRAFT include a variety of domain-specific setups, including \iac{SDR}-based testbed as a well as a ground vehicular robot for mobility-related experimentation.

The rest of this paper is structured as follows.
\cref{sec:ainur} presents the framework as well as the key concepts and technologies supporting its design and architecture.
This section also discusses briefly the assumptions made about the underlying hardware on which the framework is set to run, and we present an overview of our experimental testbed.
Next, \cref{sec:demo} describes two demonstration procedures through which we will showcase the flexibility and potential of this tool for the automatic, repeatable, end-to-end experimentation in the context of wireless testbeds.
Finally, in \cref{sec:conclusion} we summarize our contributions, discuss future directions, and conclude this paper.

In terms of future work, our current efforts are focused on the expansion and integration of Ainur with \ac{CHI}~\cite{keahey2020lessons}.
While we believe Ainur to be unique and valuable in its focus on fully-automated end-to-end wireless edge computing experimentation, integrating with the above software stacks will provide a number of features and functionalities which will greatly expand Ainur's potential.
Features like resource reservation, multi-tenant testbed access, and automated bare-metal and \ac{VM}-host instantiation will make Ainur better suited for larger scale experimentation and allow us to expand the research domain targeted by the framework.
Our ultimate goal is to eventually provide Ainur as a service running inside an OpenStack/\ac{CHI} environment, leveraging the flexibility of the platform to automate the reservation of resources, instantiation of nodes and networks, execution of experiments, and final collection of results.

\todo[inline]{Intro and conclusions from PLOS paper}

Wearable Cognitive Assistants, WCA for short, have recently started to garner attention from the research community~\cite{Ha:TowardsWearableCogAssist,Chen:EarlyImplementation}.
They represent a novel category of highly interactive and context-sensitive augmented reality applications, that aim to amplify human cognition in both day-to-day tasks and professional settings.
Their mode of operation is analogous to how GPS navigation systems guide drivers, by seamlessly providing relevant instructions and feedback relating to the current task at hand.
Note that this implies seamless interaction with the context of the user --- at no moment does the user need to trigger an update explicitly, as the application is constantly tracking the state of the target task.
An example is an IKEA Assistant~\cite{IKEAAssistant} which monitors the assembly of a piece of furniture in real time, providing timely, step-by-step feedback to guide the user toward completion.

WCA systems were originally inspired by assistive use cases for people suffering from some form of cognitive decline, either through aging or because of traumatic brain injuries~\cite{Ha:TowardsWearableCogAssist,satya2019augmenting}.
More recently, they have been applied to a broader range of use cases, including step-by-step guidance on complex assembly tasks~\cite{Chen:AnEmpiricalStudyOfLatency}.
Non-wearable augmented reality and cognitive assistance systems have already been proven to be valuable tools in the industrial workplace~\cite{funk2015caworkplace,gorecky2011cognito}.
Detethering this assistance from its current fixed location will surely make it available to many more fields.

Based on these use cases, we identify three main requirements for WCA:\@
\begin{enumerate}
    \item\label{item:pervasive} WCA systems should be available whenever the user requires them, without being tethered to a particular physical location. Assistants need to be pervasive and mobile.

    \item\label{item:interaction} Interaction with the system should be immersive and seamless, i.e.\ the assistant should be able to analyze the current context and automatically provide relevant feedback without explicit commands from the user.
    In this sense, WCA is expected to operate much like a human assistant would, by observing the performance of the user and offering guidance proactively.

    \item\label{item:lowlatency} Feedback should be ``quick'', relative to the task at hand. 
    This requirement is further strengthened by the previously mentioned ``seamless interaction'' characteristic of these systems.
    This means that users will have expectations of constant, immediate feedback as they progress through the task.
    In the case of a step-by-step task like IKEA, delayed feedback might simply confuse or distract the user.
    However, in a highly interactive task like a \emph{Ping-Pong assistant}~\cite{PingPongAssistant, Chen:EarlyImplementation} late guidance is at best a nuisance and at worst a severe handicap.
\end{enumerate}

\cref{item:pervasive} implies use of lightweight and low-power devices, preferably a wearable device that frees both hands for work.
\cref{item:interaction}, on the other hand, suggests a level of context sensitivity and proactivity that can only be provided by real-time analysis of sensor inputs such as video and audio feeds.
The compute-intensive processing suggested by \cref{item:interaction} cannot be met by the lightweight wearable devices suggested by \cref{item:pervasive}.
Only by offloading computation from a wearable device to cloud-based or edge-based infrastructure can this circle be squared.
However, offloading implies an extended end-to-end pipeline with many potential sources of queueing, transmission, and processing delays.
\cref{item:lowlatency} therefore emerges as a key concern, requiring deep understanding of the impact of end-to-end delays on WCA users.

% However, additional challenges remain in the path toward widespread adoption.
\cref{item:lowlatency} forms the base motivation for the research presented in this paper.
We still have a very limited understanding of how humans react to delays in these systems --- specifically, how changes in system responsiveness impact users.
\emph{System responsiveness} here denotes a qualitative scale ranging from \emph{high} (that is, not subject to delay or subject to negligible delay with respect to human perception) to \emph{low} (i.e.\ subject to considerable delay).
% Conversely, QoE is understood to range from \emph{good} (i.e.\ users enjoy interacting with the application) to bad (users dislike interacting with the application).
Characterizing the relationships between system responsiveness and user behavior and quality of experience is of paramount importance for the design and evaluation of these applications.
It is generally acknowledged, for instance, that a system going from a state of high responsiveness to one of low responsiveness can cause a drop in quality of experience~\cite{dabrowsky:2011:40years}. 
Furthermore, this could cause users to modify the temporal paramaters of their behavior when interacting with the system, generating a sort of \emph{feedback loop} between user and system.
A clear understanding of these relationships would allow, for instance, for the development of strategies for load balancing and optimization for large-scale deployment of WCA.\@
% Although maximizing QoE through over-engineered systems and networks would be a ``simple'' solution to this problem, in practice, the prohibitive costs of infrastructure make finding the optimal trade-off between QoE and cost a valuable endeavor.

@@

The structure of this paper is as follows.
We describe the existing body of research around time perception and the effects of delay on human performance in \cref{sec:background}.
\cref{sec:experimentaldesign} presents the experimental design, measures, and specific protocol.
Then, in \cref{sec:results} we detail the results of our experiment.
Implications for further modeling of the effects of delay are presented in \cref{sec:discussion} before finally concluding the paper in \cref{sec:conclusion}.

In this paper, we presented the results of a study on the physiological and behavioral reactions of users of WCA to delays in the application pipeline.
Our ultimate aim was to identify and categorize the ways in which humans react to low system responsiveness in step-based cognitive assistance systems.

We approached this in an experimental manner, by having participants interact with an instrumented WCA setup, and found that delay appears to affect the cognitive plan of users, preventing them from automating the task they are performing.
The results show that user interactions in a WCA slow down in the presence of delays in the application pipeline.
When system responsiveness is high, the user responds quickly; when it is low, the user slows down.
This was evidenced by an increase in task execution times, even after accounting for the artificially introduced delays, as well as accelerometer data from participants wrists.
Additionally, we found that the strength of this effect is modulated by individual differences between subjects.

These results are interesting as they open up hitherto unexplored opportunities for the design, optimization and benchmarking of WCA systems.
In this context, we believe there are two direct and important next steps to be performed.
The first of these relates to the implications for system optimization and resource allocation discussed in \cref{sec:discussion}.
We believe these need to be implemented, tested and validated in real setups.
In particular, we identify two of these implications to be prime candidates for their own experimental studies.
One, our postulation that in cases where an impaired application is close to finishing, diverting resources to it might not be the most optimal course of action; and two, the possibility that ``fair'' degradation of system responsiveness across applications may be, in some cases, undesirable.
Both of these questions could be answered with straightforward setups.

The second step corresponds to the extension of existing tools for WCA benchmarking with the findings presented in this paper.
These tools are of great usefulness for the study of WCA systems, as they allow for automated large-scale testing without having to resort to human users.
However, they are still somewhat simplistic and unrealistic in their workload generation schemes.
Incorporating the results presented here would allow for much more realistic workloads.
For instance, our findings relating to the effects of delay on execution times could be directly adapted to modulate timings in the input stream.
Another example would be employing the results linking neuroticism to a heightened sensibility for delays to provide a ``tuning knob'' for the user models in these workload generation schemes.
Extending and perfecting these tools will allow for much more realistic benchmarking and testing of WCA systems, providing data of significantly better quality and ultimately leading to faster improvement, optimization and adoption of these systems.

\todo[inline]{Intro and conclusions from EdgeDroid 2.0}

\acf{WCA} applications are a novel category of wearable, edge-native applications aiming to amplify human cognition in both daily activities and professional settings.
These systems aim to seamlessly integrate into the day-to-day of users, leveraging compute-intensive algorithms to analyze user and environment information and provide real-time, context-aware information and feedback.
\ac{WCA} applications originally emerged as assistive use-cases for individuals suffering from cognitive decline due to aging or traumatic brain injuries~\cite{satyanarayanan2009case,Ha2014towards,Satya2019augmenting}, and have since expanded to a greater range of use cases.
In particular, following the success of non-wearable \ac{XR} and cognitive assistance in industrial settings~\cite{Funk2015Cognitive,Wang2022Comprehensive}, there is increasing interest in the research community in the application of \ac{WCA} for step-by-step assistance for complex assembly tasks~\cite{Chen2017Empirical,belletier2021wearable}.

A defining characteristic of these applications is their lack of reliance on intentional user inputs to trigger responses.
They are intended to operate as autonomous guides, much akin to how \ac{GPS} systems guide drivers, tracking their progress and providing feedback and instructions at appropriate times.
This context-sensitivity and proactivity in providing user feedback translates into a reliance on high-dimensional, complex, unstructured inputs, such as real-time video feeds, which require intensive compute capabilities to process.
This also translates into latency-sensitivity, as with any \ac{AR} application.
Delays and jitter can be jarring to the user, causing them discomfort, leading to make mistakes and potentially even to abandonment of the application altogether.
On the other hand, as their name suggests, these systems are by design \emph{wearable}, which implies the use of lightweight, battery-powered, low energy consumption devices.

>>>>

We summarize the seven key contributions of this thesis below:

@@

\todo[inline]{
    \cref{paper:EdgeDroid1.0} introduces our methodology, and validates its use in studying WCA.
    Focus less on EdgeDroid 1.0 and more on the methodology, although discuss that EdgeDroid 1.0 is an initial approach to a tool we complete with EdgeDroid 2.0.
}

\begin{figure}
    \centering%
    \includegraphics[width=.85\columnwidth]{publications/2019EdgeDroid/plots/comparison/nofonts/rtt_fb_vs_nofb}%
    \caption{Comparison of round-trip-times for inputs that triggered a state transition in the task model versus inputs that did not.}%
    \label{fig:comparison:rtt}%
\end{figure}%
\begin{figure}
    \centering%
    \includegraphics[width=.85\columnwidth]{publications/2019EdgeDroid/plots/comparison/nofonts/box_feedback}%
    \caption{Distribution of latency across system components for inputs that triggered a state transition in the task model.}%
    \label{fig:comparison:feedback}%
\end{figure}%
\begin{figure}%
    \centering%
    \includegraphics[width=.85\columnwidth]{publications/2019EdgeDroid/plots/comparison/nofonts/box_taskstep}%
    \caption{Round-trip-times for input-feedback cycles associated with state transitions in the internal task model for a single client connected over an optimal wireless link.}%
    \label{fig:comparison:tasksteps}
\end{figure}%

@@

\Cref{fig:comparison:rtt} presents a comparison of the total measured round-trip-times (RTTs) both for inputs which caused a state transition and inputs that did not.
Next, \cref{fig:comparison:feedback} shows a comparison of the distribution of latencies across system components for inputs that caused an internal state change in the task model of the application.
We differentiate according to the three main components contributing to latency, namely \emph{uplink} and \emph{downlink transmissions}, and \emph{backend processing}.
Finally, \cref{fig:comparison:tasksteps} depicts the distribution of RTTs for each transition in the internal task model for a single client connected over an optimal wireless link.
These metrics were calculated by recording the measured input-feedback cycle delay corresponding to a change in state within the application.
Thus, for instance, the measurements located at column ``3 to 4'' in the figure correspond to the aggregated round-trip-times for every input-feedback cycle corresponding to a change from state 3 to state 4 within the application task model, for the 100 repetitions of the scenario.

In the following discussion we will refer to inputs which triggered a transition in the task model as \emph{feedback-rich inputs} and those that did not as \emph{feedback-less inputs}.

For the analysis of these results, we will take into consideration the bound of \SI{600}{\milli\second} response time for step-by-step task-guidance derived by the authors of~\cite{Chen:AnEmpiricalStudyOfLatency}.
This bound marks the point after which further delays in the delivery of feedback to the user start to negatively affect user experience, and allows for a straightforward evaluation of the responsiveness of the system.

We will begin our analysis of the experiment results with \Cref{fig:comparison:rtt}.
These results present a stark contrast in the round-trip-times for inputs which cause a state transition versus inputs that do not, with RTTs for the former being up to an order of magnitude greater.
It's worth mentioning though that responses to feedback-less inputs are invisible to the user, and are just included here as a sort of baseline to compare feedback-rich round-trip times with.

We can identify a pair of interesting effects in the scaling of the task-guidance WCA application.
One, scaling behavior for the application seems to be linear with respect to the number of clients.
Two, in the case of the impaired WiFi, the effect on the feedback-rich inputs is very pronounced, with the average of the RTTs for these inputs being over the previously discussed bound of \SI{600}{\milli\second}.

It is worth noting that already at just 10 clients the response times for feedback-rich inputs are very close to the bound.
Looking at this through the lens of a an application developer, it could hint at a need for optimization of the later parts of the application pipeline, since RTTs for inputs which are discarded in the detection stage of the pipeline (i.e.\ feedback-less inputs) are still well below \SI{200}{\milli\second}.

EdgeDroid 1.0 allows researchers to zoom into specific components of the application feedback loop, as exemplified by \cref{fig:comparison:feedback}.
From this figure it is clear that the main component which contributes to latency in the optimal case is the backend processing, further lending credibility to our previous comment on the need for optimization.
Nevertheless, when the link quality decreases, the delays on the downlink start to overshadow the delays on the processing.
Here, the downlink time sometimes almost reaches the ideal bound by itself.
A system designer might then conclude from this that in order to be able to scale the application, their focus needs to be on improving the quality of the wireless link before increasing the processing power on the backend.

Finally, EdgeDroid 1.0 allows even more insights to be gained by homing in to individual steps in a task-guidance WCA. Consider \cref{fig:comparison:tasksteps}.
The figure shows clears spikes in latencies at the transitions from task state 3 to 4 and from task state 4 to 5, which could indicate to the application developer that these specific transitions are ripe for optimization.

\todo[inline]{%
    Introduce as follow-up to methodology discussion.
    Natural implication of that is that we need a more detailed model of human behavior.
    EdgeDroid 1.0 was an initial, very coarse approximation.
    Here we present our final, polished model based on empirical data.
}

\todo[inline]{PLOS paper describes the acquisition of the necessary data and insights for the models.}

We still have a very limited understanding of how humans react to delays in these systems --- specifically, how changes in system responsiveness impact users.
\emph{System responsiveness} here denotes a qualitative scale ranging from \emph{high} (that is, not subject to delay or subject to negligible delay with respect to human perception) to \emph{low} (i.e.\ subject to considerable delay).
% Conversely, QoE is understood to range from \emph{good} (i.e.\ users enjoy interacting with the application) to bad (users dislike interacting with the application).
Characterizing the relationships between system responsiveness and user behavior and quality of experience is of paramount importance for the design and evaluation of these applications.
It is generally acknowledged, for instance, that a system going from a state of high responsiveness to one of low responsiveness can cause a drop in quality of experience~\cite{dabrowsky:2011:40years}. 
Furthermore, this could cause users to modify the temporal paramaters of their behavior when interacting with the system, generating a sort of \emph{feedback loop} between user and system.
A clear understanding of these relationships would allow, for instance, for the development of strategies for load balancing and optimization for large-scale deployment of WCA.\@

@@

We start our discussion with the main results of our experimentation presented in the previous section:

\begin{itemize}
  \item Firstly, and perhaps most importantly, we find that a system slow-down induces an \textit{additional} behavioral slow-down.
  That is, as system responsiveness decreases, our data indicates that users significantly slow-down in their execution of the task.
  This slow-down scales with the decrease in responsiveness; compared to the no-delay case, participants were on average \SI{12}{\percent} slower at \SI{1.65}{\second} delay and \SI{26}{\percent} at \SI{3.0}{\second} delay.
  Moreover, there is a temporal component to this effect; users become progressively slower the more time passes with reduced system responsiveness.

  \item Secondly, we find that the effects of behavioral slow-down due to impaired system responsiveness \emph{remain} for at least a few steps after system responsiveness improves.
  This is evidenced by the longer per-step-execution times of the first four steps of blocks immediately following a high-delay block, as pictured in \cref{fig:exectime:transition}.
  The question of whether any lingering effect can be measured after these four steps remains open.

  \item Thirdly, we evidence a speed-up in execution time over a series of steps; that is, subjects get faster at performing steps as the task progresses.
  However, the strength of this effect decreases as delay increases.
  Whereas for blocks without delay users performed the last four steps of a 12-step block on average \SI{36}{\percent} faster than the first four, at the maximum delay this effect practically disappears.

  \item Fourthly, in terms of inter-subject differences, PCA revealed three main factors governing users' response to delay. 
  The first factor represents sensitivity to delay as moderated by the ``Big Five'' personality trait of neuroticism and both measures of immersion: focus and involvement.
  Factor two and three represent dedication to the task as opposed to delay intolerance and reflect variables related to attentiveness, respectively.
  In simple terms, these results suggest that the effects of delay are most potent in individuals who are sensitive and involved in the task.
  The findings appear selective to cognitive assistance tasks like the present ones, inasmuch as the same measures did not correlate with outcomes in other computer-intensive environments such as immersive VR~\cite{quesnel2018you}.
  These correlations are also consistent with previous findings indicating that individuals scoring high in neuroticism tend to be intolerant to delayed reward~\cite{hirsh2008delay}.
\end{itemize}

A central question therefore arises: to which physio- and psychological mechanisms can these findings, most importantly the substantial slow-down in task execution, be attributed?

In \cref{sec:background}, we initially considered the possibility that delays might produce negative emotional reactions.
These could in turn elicit generalized arousal.
We also postulated that adapting to delay might progressively deplete cognitive resources in users.
However, the present data provide relatively little support for these alternatives, in that the predicted measures did not produce the expected statistical trends.
Specifically, physiological measures of GSR and HR failed to show evidence of differential arousal under long vs.\ short delays, and speed-induced errors and non-completions predicted by resource depletion were not observed.
The acceleration data further did not indicate that extended delay significantly increases erratic movement.
% physiological measures of GSR and HR failed to show evidence of differential arousal under long vs.\ short delays, and speed-induced errors and non-completions predicted by resource depletion were not observed.
% The acceleration data further do not indicate that extended delay increases erratic movement.
To the contrary, the data suggest this effect results from a delay in movement after an instruction is introduced.
That is, users fail to capitalize on the new information as quickly as they could.
Thus, contrary to our preliminary postulations, behavioral effects seem to arise from impaired cognitive control mechanisms, and not from emotion or resource depletion.
We hypothesize that the effects of feedback latency can best be understood as changes in the use of a cognitive plan.
As was described in \cref{sec:experimentaldesign,fig:lego:hierarchical}, complex cognitive and motor tasks have been modeled as the unfolding of a hierarchy of command, from high-level plans to physical output.
Long system latencies, we propose, disrupt the automating of such a plan, instead relegating it to attention-based control at the step-by-step level that is easily diverted.
This also provides a possible explanation for the lingering effects of delay after an acceptable system responsiveness is restored, as users needs time re-adjust and re-automate their cognitive plans.

As to the applicability of our findings to other applications, it must be noted that these results pertain to a specific class of applications, namely step-based task-guidance WCA.\@
However, we would expect our findings to extend to similar applications, as long as they follow the same pattern of seamless interaction --- i.e.\ such that the user does not need explicitly interact with the application to advance the state.

The results here presented provide a number of possible implications for WCA system design and optimization, both for single and multi-application flows.

\begin{itemize}

  \item Due to the behavioral slow-down in users, even short-term reductions in responsiveness will lead to significantly extended application lifetimes.
  This has direct implications for resource and power consumption.

  \item The fact that the adverse effects of delay on users do not immediately disappear as the system returns to a high-responsive state could have unconventional consequences for resource allocation.
  This is of particular importance, for instance, for cases where the user may be able to finish the task before these effects subside.
  In such cases, the limited potential gains might not justify diverting valuable system resources to the impaired application.

  \item In multi-user environments, the time dependency of user slow-down effects mean that fair degradation of system responsiveness across applications may not ultimately be beneficial to the system as a whole.
  Take for instance two applications on the same system which negatively interfere with each other. 
  The longer they interfere with each other, the longer their respective lifetimes are going to be, which in turn causes them to interfere even longer, potentially entering a positive feedback loop. 
  In such a case, prioritizing one over the other rather than trying to improve responsiveness for both might lead to resources being freed up faster system-wide.

  \item Based on our findings relating individual differences between users and their sensitivity to delays, it might also be possible to extrapolate user characteristics from measured execution times.
  This could prove a valuable tool for load balancing, for instance by prioritizing resource allocation to users with a higher sensitivity to system-state degradation.
  However, this remains an open challenge.
